---
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
author: Irem B. Gunduz
date: "`r Sys.Date()`"
title: |
  | Allele-specific chromatin accessibility
  | using ChrAccR
vignette: >
  %\VignetteIndexEntry{allelespec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = ">",
  fig.width = 7,
  fig.height = 4.5
)
runDemo     <- TRUE
examplePath <- system.file("extdata", "dsASC_example.rds", package = "ChrAccR")
useExample  <- runDemo && nzchar(examplePath)
useSim      <- runDemo && !nzchar(examplePath)
```

# Introduction

A heterozygous site carries two alleles that share one nucleus, one set of trans-acting factors and one library preparation. Anything that differs between the reads carrying the reference allele and the reads carrying the alternative allele therefore has to come from the sequence itself. This is what makes allele-specific chromatin accessibility (ASC) a better-controlled measurement than comparing accessibility between individuals: the second allele is the control, and it is a matched one.

`AlleleSpeC` is the ASC layer of `ChrAccR`. It takes WASP-corrected BAM files and donor genotypes, counts reference and alternative reads at every heterozygous site inside consensus peaks, tests each site for imbalance, and asks whether the imbalance follows the allele a transcription factor is predicted to bind. The data structure carrying all of this is `DsASC`, a `DsAcc` subclass holding two aligned count matrices (`ref` and `alt`, sites x samples) alongside the allele-agnostic peak accessibility matrix, so that allelic and total signal never drift apart.

This vignette walks the pipeline end to end. The heavy stages need cluster resources and tens of gigabytes of BAM files, so they are shown but not executed. Everything downstream of counting runs on a small example object shipped with the package, cut from a real run by `inst/scripts/make_vignette_data.R`. If that object has not been built yet the vignette simulates an equivalent dataset instead, so it always knits. The last section covers what stays out of the repository, and why.

# The shape of the pipeline

| Stage | Runs in | Entry point | Produces |
|---|---|---|---|
| Genotyping | shell (GATK) | `01_call_variants.sh` | `{donor}_heterozygous.vcf.gz` |
| Mapping-bias correction | shell (WASP) | `03_wasp.R` | `{sample}_wasp.bam` |
| Shared site list | R | `buildMasterSNPs()` | master `GRanges` with donor membership |
| Allele counting | R | `DsASC.gatk()` | one `DsASC` per cell type |
| Peak restriction | R | `annotateASCSites()`, `filterASCByPeaks()`, `setAccessibility()` | `DsASC` limited to open chromatin |
| Genotype and bias cleanup | R | `ascDropHomozygous()`, `ascFilterRegions()`, `ascPoolReplicates()` | cleaned count matrices |
| Imbalance testing | R | `calcASCStatistics()`, `filterForRecurrence()` | per site-sample statistics |
| Cross-cell-type sharing | R | `ascAggregate()`, `ascSharing()`, `estimateSharedImbalance()` | sharing estimates |
| Motif disruption | R | `motifmatchr` scoring | per-site PWM delta table |
| Binding preference | R | per-site PWM preference bins | reference fraction by predicted preference, per condition |

The first two stages are the expensive ones, and they are also the ones that decide whether the result is real. Everything after them is comparatively cheap, and the package functions assume both have been done properly.

# Requirements

Beyond the standard `ChrAccR` dependencies, the ASC layer draws on a few additional packages. Genotype reading and pileup counting need `VariantAnnotation` and `Rsamtools`; disk-backed counting needs `HDF5Array` and `DelayedArray`; the sharing estimate needs `ashr`; the motif stage needs `JASPAR2020`, `TFBSTools`, `motifmatchr` and a `BSgenome` package for the assembly you work in.

```{r, eval = FALSE}
BiocManager::install(c("VariantAnnotation", "Rsamtools", "HDF5Array", "DelayedArray",
                       "JASPAR2020", "TFBSTools", "motifmatchr",
                       "BSgenome.Hsapiens.UCSC.hg38"))
install.packages("ashr")
```

```{r, message = FALSE, warning = FALSE}
library(ChrAccR)
library(GenomicRanges)
library(data.table)
library(ggplot2)
theme_set(theme_bw(base_size = 11))
```

# Before R: genotypes and bias correction

Two things have to be true before a single allele is counted, and no amount of downstream statistics can repair either of them.

The first is that the heterozygous sites really are heterozygous. Genotypes are called per donor, jointly over all of that donor's libraries, so that a site is supported by the donor's full read depth rather than by one shallow sample. Only biallelic SNVs called heterozygous are kept, and the result is written one file per donor.

```{r, eval = FALSE}
system("bash prep/01_call_variants.sh tables/sampleAnnot_filtered.csv")
```

The second is that reads carrying the alternative allele are not quietly lost during alignment. A read that differs from the reference maps slightly worse than one that matches it, which pushes every site toward the reference allele and manufactures ASC out of nothing. WASP removes this by re-mapping each overlapping read with its alleles swapped and discarding any read that does not return to the same position.

```{r, eval = FALSE}
system("Rscript prep/03_wasp.R")
system("Rscript prep/04_wasp_audit.R")
```

The audit step is worth running even when the pipeline reports success. It measures the pooled reference fraction per sample at that donor's heterozygous sites; a corrected library sits near 0.5, and anything drifting above roughly 0.55 is still biased and will contaminate every result that follows.

The sample annotation that carries the rest of the pipeline needs `sampleId`, `bamFilename` and `donor` at minimum, with `cellType` and `stimulus` added for grouping and contrasts. Row names must be the sample identifiers.

```{r, eval = FALSE}
meta <- fread("tables/sampleAnnot_filtered.csv")
meta[, donor := as.character(donor)]
meta[, bamFilename := file.path(waspDir, paste0(sampleId, "_wasp.bam"))]
meta <- meta[file.exists(bamFilename)]
```

# One site list for every donor

Each donor has their own heterozygous sites, so counting each cell type against its own site list would produce objects whose rows cannot be compared. `buildMasterSNPs()` reads every donor VCF once and returns the sorted union of biallelic heterozygous SNPs, carrying with each site the set of donors heterozygous there in a `donorMembership` column.

```{r, eval = FALSE}
masterGr <- buildMasterSNPs(donors = unique(meta$donor), vcfDir = vcfDir, genome = "hg38")
saveRDS(masterGr, file.path(ascDir, "masterSNPs.rds"))
```

Built once and reused, this list keeps every per-cell-type object row-aligned, which is the precondition for later asking whether a site imbalanced in T cells is also imbalanced in B cells. The membership column matters just as much: it is what lets a sample be counted only where its own donor is heterozygous, and be marked missing everywhere else rather than silently treated as a balanced site.

# Counting alleles

`DsASC.gatk()` performs the pileup. Sites outside a sample's donor het set receive `NA` rather than zero, so a homozygous genotype can never be mistaken for a covered site with no imbalance. Genuine zeros inside the donor's het set are kept as zeros. Pileup stringency comes from the package configuration.

```{r, eval = FALSE}
setConfigElement("minMapq", 20L)
setConfigElement("minBaseq", 20L)
setConfigElement("maxDepth", 5000L)

dsAsc <- DsASC.gatk(
  sampleAnnot = as.data.frame(meta[cellType == "Tcell",
                                   .(sampleId, bamFilename, donor, cellType, stimulus)]),
  vcfDir      = vcfDir,
  genome      = "hg38",
  diskDump    = TRUE,
  masterGr    = masterGr
)
```

With `diskDump = TRUE` the two count matrices are written straight into HDF5 sinks, one sample column at a time, so memory stays flat regardless of how many million sites the master list holds. Realize them back into memory before serializing the object, otherwise the saved `.rds` points at an HDF5 file that will not travel with it.

Because each cell type is independent given the shared master list, the counting stage parallelizes cleanly as a job array: one serial R process per cell type, all writing into the same directory. `src/01_asc_counting_ChrAccR.R` is written for exactly this, taking either `build_master` or a cell type name as its single argument.

# Restricting to open chromatin

A heterozygous site is only informative where there is chromatin signal for it to be allele-specific about. The peak set comes from the ordinary `ChrAccR` ATAC workflow, and the same object supplies the allele-agnostic accessibility matrix that is attached alongside the allele counts.

```{r, eval = FALSE}
dsAtac  <- loadDsAcc(atacDir)
peakGr  <- getCoord(dsAtac, "cons_peaks")
peakMat <- getCounts(dsAtac, "cons_peaks")
names(peakGr)     <- paste0("peak_", seq_along(peakGr))
rownames(peakMat) <- names(peakGr)

dsAsc <- filterChroms(dsAsc, remove = c("chrX", "chrY", "chrM"))
dsAsc <- filterLowCovg(dsAsc, thresh = 5, reqSamples = 1)
dsAsc <- annotateASCSites(dsAsc, peakGr)
dsAsc <- filterASCByPeaks(dsAsc)
dsAsc <- setAccessibility(dsAsc, peakMat[, getSamples(dsAsc)], peakGr)
```

Sex chromosomes go first because allelic balance is not defined there, and the coverage filter runs before peak annotation because it is the cheaper of the two and removes most of the rows.

# The example dataset

Everything from here on runs on four objects: a `DsASC`, the consensus peaks and accessibility matrix that go with it, and a table of per-site motif deltas. The packaged example holds all four, cut down from a full run to a handful of cell types from different lineages, resting and stimulated samples only, one chromosome and a few thousand covered sites. Thymic populations are excluded: they are the shallowest libraries in the dataset, several produce empty objects at the counting stage, and none of them survive coverage filtering in numbers that would demonstrate anything.

`inst/scripts/make_vignette_data.R` is what produces that file, and it reports the size of what it wrote so the example cannot quietly outgrow the package. Rerun it against your own `asc_counts` directory to regenerate the example, adjusting the cell types, chromosome and site count at the top of the script.

```{r example-load, eval = useExample}
example   <- readRDS(examplePath)
dsDemo    <- example$ds
peakGr    <- example$peakGr
peakMat   <- example$peakMat
deltaDemo <- example$delta
dsDemo
```

If the example has not been built, the block below stands in for it, with the same structure a real object has after counting: three donors, two cell types, two stimulation states, two libraries each, and 400 heterozygous sites. Roughly one site in six is given a genuine allelic imbalance, a handful are made to look homozygous in one donor so the cleanup step has something to find, and the direction of each real imbalance is tied to a per-site motif effect that the transcription-factor section will recover.

```{r sim, eval = useSim}
set.seed(42)

nSites   <- 400L
donors   <- c("D1", "D2", "D3")

sampleAnnot <- expand.grid(
  replicate = c(1, 2), stimulus = c("U", "S"),
  cellType  = c("Tcell", "Bcell"), donor = donors,
  stringsAsFactors = FALSE
)
sampleAnnot$sampleId <- with(sampleAnnot,
  paste(donor, cellType, stimulus, replicate, sep = "_"))
sampleAnnot <- sampleAnnot[, c("sampleId", "donor", "cellType", "stimulus", "replicate")]
rownames(sampleAnnot) <- sampleAnnot$sampleId
nSamples <- nrow(sampleAnnot)

pos    <- sort(sample(seq(1e6, 5e6), nSites))
refAll <- sample(c("A", "C", "G", "T"), nSites, replace = TRUE)
altAll <- unname(vapply(refAll, function(b) sample(setdiff(c("A", "C", "G", "T"), b), 1L), ""))
snpId  <- paste0("chr1:", pos, "_", refAll, "_", altAll)

snpGr <- GRanges("chr1", IRanges(pos, pos), REF = refAll, ALT = altAll, snpId = snpId)
names(snpGr) <- snpId

siteDelta <- rnorm(nSites, 0, 1.5)
names(siteDelta) <- snpId

isImbalanced <- rep(FALSE, nSites)
isImbalanced[sample(nSites, 70L)] <- TRUE
altFrac <- ifelse(isImbalanced, 0.5 + 0.22 * sign(siteDelta), 0.5)

depth  <- matrix(rpois(nSites * nSamples, lambda = 25L), nrow = nSites,
                 dimnames = list(snpId, sampleAnnot$sampleId))
altMat <- matrix(rbinom(nSites * nSamples, size = as.vector(depth), prob = altFrac),
                 nrow = nSites, dimnames = dimnames(depth))
refMat <- depth - altMat

leakSites   <- sample(nSites, 20L)
leakSamples <- sampleAnnot$sampleId[sampleAnnot$donor == "D1"]
refMat[leakSites, leakSamples] <- refMat[leakSites, leakSamples] + altMat[leakSites, leakSamples]
altMat[leakSites, leakSamples] <- 0L

inPeak  <- sort(sample(nSites, round(0.8 * nSites)))
peakGr  <- GRanges("chr1", IRanges(pos[inPeak] - 200L, pos[inPeak] + 200L))
names(peakGr) <- paste0("peak_", seq_along(peakGr))
peakMat <- matrix(rpois(length(peakGr) * nSamples, lambda = 40L), nrow = length(peakGr),
                  dimnames = list(names(peakGr), sampleAnnot$sampleId))

dsDemo <- DsASC(sampleAnnot, genome = "hg38")
dsDemo@coord  <- list(snps = snpGr)
dsDemo@counts <- list(ref = refMat, alt = altMat)

deltaDemo <- rbindlist(list(
  data.table(snpId = snpId, tf = "JUNB",  delta = as.numeric(siteDelta)),
  data.table(snpId = snpId, tf = "BACH1", delta = as.numeric(0.5 * siteDelta + rnorm(nSites))),
  data.table(snpId = snpId, tf = "FOSL1", delta = as.numeric(sample(siteDelta)))
))
dsDemo
```

That assembly is exactly what `DsASC.gatk()` does internally: an empty `DsASC` shell from the sample annotation, the master sites into `coord`, the two matrices into `counts`.

Either way, the filtering chain applies from here on and the accessibility matrix is attached. On the packaged example the peak restriction is close to a no-op, because the counting stage already applied it; running it again costs nothing and keeps the two paths identical.

```{r filter, eval = runDemo, message = FALSE}
dsDemo <- filterChroms(dsDemo, remove = c("chrX", "chrY", "chrM"))
dsDemo <- filterLowCovg(dsDemo, thresh = 5, reqSamples = 1)

if (!is.null(peakGr)) {
  dsDemo <- annotateASCSites(dsDemo, peakGr)
  dsDemo <- filterASCByPeaks(dsDemo)
}
if (!is.null(peakMat) && all(getSamples(dsDemo) %in% colnames(peakMat))) {
  dsDemo <- setAccessibility(dsDemo, peakMat, peakGr)
}
dsDemo
```

The accessors return plain matrices, and `getAllelicBalance()` gives the alternative-allele fraction with a coverage floor applied, which is the quantity every plot in this vignette is ultimately built on.

```{r accessors, eval = runDemo}
dim(getRefCounts(dsDemo))
balance <- getAllelicBalance(dsDemo, minCoverage = 10)
summary(as.vector(balance))
```

# Cleaning before testing

Three corrections belong between counting and testing, and all three operate on the count matrices rather than on the object, because they change what a column means.

Mis-genotyped sites come first. A site called heterozygous but actually homozygous produces a perfect, highly significant imbalance, and there will be many of them. `ascDropHomozygous()` pools each donor's samples and zeroes the sites where the pooled minor allele is either absent or present only at an error-level fraction, so they fall out of the coverage filter instead of topping the results table.

```{r drophom, eval = runDemo, message = FALSE}
annotDemo <- getSampleAnnot(dsDemo)
cleaned <- ascDropHomozygous(getRefCounts(dsDemo), getAltCounts(dsDemo), annotDemo,
                             minMinor = 2L, minMinorFrac = 0.05, minTotal = 10L)
```

Region masking comes next. Repetitive and low-mappability regions are where WASP fails and where false heterozygous calls accumulate, and both push the result toward the reference allele. `ascFilterRegions()` takes the site identifiers and an ENCODE blacklist and returns the ones worth keeping.

```{r, eval = FALSE}
keepIds <- ascFilterRegions(rownames(cleaned$ref), blacklistGr)
cleaned$ref <- cleaned$ref[keepIds, ]
cleaned$alt <- cleaned$alt[keepIds, ]
```

Finally, the unit of analysis. A binomial test on a technical replicate uses half the reads and loses most of the power that the second half would have given it; the biological sample, not the library, is what should be tested. `ascPoolReplicates()` sums libraries sharing a donor, cell type and condition and returns a matching annotation.

```{r pool, eval = runDemo, message = FALSE}
pooled <- ascPoolReplicates(cleaned$ref, cleaned$alt, annotDemo,
                            groupCols = c("cellType", "stimulus", "donor"))
dim(pooled$ref)
head(pooled$annot, 3)
```

# Testing for allelic imbalance

Each site-sample pair with enough coverage is tested against a fixed null of 50:50 with a two-sided binomial test, and the p-values are corrected within each sample. The null stays at 0.5 deliberately: WASP is responsible for removing mapping bias, and re-centring the null on the observed allele fraction would calibrate it against the very signal being detected. The `minAllele` guard requires reads on both alleles, which removes the low-coverage calls whose only route to significance is a minor allele of exactly zero.

```{r stats, eval = runDemo, message = FALSE}
ascStats <- calcASCStatistics(dsDemo, minCoverage = 10, minAllele = 2)
ascStats[, donor := annotDemo$donor[match(sampleId, annotDemo$sampleId)]]
ascStats[order(fdr)][1:5, .(snpId, sampleId, donor, ref, alt, total, log2FC, fdr)]
```

The three standard views of that table are a per-site forest plot, a volcano and a genome-position view.

```{r plots, eval = runDemo, message = FALSE, fig.height = 5}
topSites <- as.character(ascStats[order(fdr), unique(snpId)])[1:3]
plotASCBalance(dsDemo, topSites, colorBy = "stimulus", orderBy = "cellType",
               sigStats = ascStats, sigCut = 0.1)
```

```{r volcano, eval = runDemo, message = FALSE}
plotASCVolcano(dsDemo, stats = ascStats, sigCut = 0.1)
```

```{r manhattan, eval = runDemo, message = FALSE}
plotASCManhattan(dsDemo, stats = ascStats, sigCut = 0.1)
```

Before reading anything into the site list, look at where the calls came from. A donor with much deeper libraries will contribute more significant sites simply because it has more power, and a donor with a bad genotype batch will contribute more because its false heterozygotes all look perfectly imbalanced. Neither is visible in a volcano, but both are obvious as soon as the calls are counted per donor.

```{r donorbar, eval = runDemo, message = FALSE, fig.height = 3.5}
sigPerDonor <- ascStats[fdr < 0.1, .(nSites = uniqueN(snpId)), by = donor]

ggplot(sigPerDonor, aes(x = reorder(donor, -nSites), y = nSites)) +
  geom_col(fill = "#34495E", width = 0.65) +
  labs(x = "Donor", y = "Significant ASC sites (FDR < 0.1)",
       title = "ASC calls contributed by each donor")
```

A roughly even set of bars is what you want. One bar several times the height of the rest is a flag to go back to that donor's coverage and genotype QC before trusting sites that only it supports.

A single significant call in a single donor is weak evidence for the same reason. Recurrence across donors is the stronger claim, and `filterForRecurrence()` counts how many donors independently support each site. It expects a `donor` column, which was attached above, and a normalised effect-size column.

```{r recurrence, eval = runDemo, message = FALSE}
recurrent <- filterForRecurrence(ascStats, minDonors = 2, fdrCutoff = 0.1)
head(recurrent, 5)
```

# Sharing across cell types

Because every cell type was counted against the same master list, per-cell-type objects can be read back and stacked into one aligned pair of matrices.

```{r, eval = FALSE}
merged <- mergeDsASCArray(ascDir)
```

The sharing question is asymmetric: take the sites significant in cell type A, and ask what happens to them in cell type B. A site can be imbalanced in the same direction, balanced, or simply not accessible enough to say. `ascSharing()` separates those three outcomes, using `ashr` to estimate the proportion with a genuinely non-zero effect in B rather than counting significance calls twice.

```{r, eval = FALSE}
cellsPresent <- unique(pooled$annot$cellType)
samplesA <- pooled$annot[cellType == cellsPresent[1], sampleId]
samplesB <- pooled$annot[cellType == cellsPresent[2], sampleId]

aggA <- ascAggregate(pooled$ref, pooled$alt, samplesA)
aggB <- ascAggregate(pooled$ref, pooled$alt, samplesB)

sharing <- ascSharing(aggA$ref, aggA$alt, aggB$ref, aggB$alt,
                      fdrA = 0.01, minReadsB = 4)
sharing
```

`ascSitePosterior()` gives the per-site posterior and FDR behind that estimate, `ascStratum()` labels a pair of groups by whether they share lineage, condition, both or neither, and `estimateSharedImbalance()` runs the same logic directly on two `DsASC` objects when the matrices have not been merged by hand.

```{r, eval = FALSE}
shared <- estimateSharedImbalance(dsTcell, dsBcell, fdrCutoff = 0.01)
shared$sharing_estimate
```

# From sites to motifs

The link between a variant and a transcription factor is the change in predicted binding it causes. For each site, a window the width of the widest motif is taken from the genome, the centre base is substituted to give reference and alternative versions, both are scored with every PWM, and the difference is recorded as `delta = score(ALT) - score(REF)`. A positive delta means the alternative allele is the predicted stronger binder.

```{r, eval = FALSE}
library(JASPAR2020); library(TFBSTools); library(motifmatchr)
library(BSgenome.Hsapiens.UCSC.hg38)

pfms  <- getMatrixSet(JASPAR2020, list(species = 9606, collection = "CORE"))
flank <- max(vapply(pfms, function(p) ncol(as.matrix(p)), 1L))

sites <- coordDt[snpId %in% sigSnps]
gr    <- GRanges(sites$chrom, IRanges(sites$pos - flank, sites$pos + flank))
seqRef <- getSeq(BSgenome.Hsapiens.UCSC.hg38, gr)
seqAlt <- replaceLetterAt(
  seqRef,
  at = matrix(seq_len(width(seqRef)[1]) == (flank + 1L),
              nrow = length(seqRef), ncol = width(seqRef)[1], byrow = TRUE),
  letter = sites$ALT)

scoreSet <- function(s) as.matrix(motifScores(
  matchMotifs(pfms, s, out = "scores", bg = "even", p.cutoff = 1)))
delta <- scoreSet(seqAlt) - scoreSet(seqRef)
colnames(delta) <- vapply(pfms, name, "")

deltaDt <- rbindlist(lapply(colnames(delta), function(tfName)
  data.table(snpId = sites$snpId, tf = tfName, delta = delta[, tfName])))
```

Scoring is restricted to the ASC-significant sites, which is what makes a genome-wide motif set tractable; scoring every site in every peak costs far more and adds nothing, since sites with no imbalance carry no allelic information.

`deltaDemo` already holds a table of this shape. In the packaged example it is the real output of the code above, computed once over the example sites by the data script. In the simulated fallback it is three synthetic profiles, one aligned with the injected imbalance, one partially aligned and one shuffled, so the section below has both a signal and a negative control.

```{r deltaShow, eval = runDemo}
deltaDemo[, .(nSites = .N, meanAbsDelta = round(mean(abs(delta)), 2)), by = tf]
```

Each site is then labelled by which allele the motif prefers. Sites whose delta clears the threshold in either direction are called for that allele; everything in between is "no preference" and acts as the internal control. The driver script uses a threshold of 1.0 on the PWM score; it is relaxed here because the example carries a few thousand sites rather than a genome's worth.

```{r preference, eval = runDemo}
prefDelta <- 0.5

prefDt <- data.table::copy(deltaDemo)
prefDt[, preference := fifelse(delta >  prefDelta, "Alternative",
                       fifelse(delta < -prefDelta, "Reference", "No preference"))]
prefDt[, .N, by = .(tf, preference)][order(tf, preference)][1:6]
```

# Binding preference and allelic accessibility

This is the test that ties the two halves together. If a variant really disrupts a factor's binding site, the allele the motif prefers should be the more accessible one, and the effect should be larger when the factor is active. Pooling reads by condition and plotting the reference-allele fraction against the predicted preference asks exactly that, with the "no preference" sites sitting at 0.5 to show the axis is not simply skewed.

```{r fracByCondition, eval = runDemo}
condFrac <- function(cond, label) {
  sids <- intersect(pooled$annot[stimulus == cond, sampleId], colnames(pooled$ref))
  r <- rowSums(pooled$ref[, sids, drop = FALSE])
  a <- rowSums(pooled$alt[, sids, drop = FALSE])
  data.table(snpId = rownames(pooled$ref), refFrac = r / (r + a),
             total = r + a, condition = label)
}

fracDt <- rbindlist(list(condFrac("U", "Resting"), condFrac("S", "Stimulated")))
fracDt <- fracDt[total >= 4]

plotDt <- merge(fracDt, prefDt, by = "snpId", allow.cartesian = TRUE)
plotDt[, preference := factor(preference,
        levels = c("Alternative", "No preference", "Reference"))]
plotDt[, condition := factor(condition, levels = c("Resting", "Stimulated"))]
```

```{r tfboxplot, eval = runDemo, fig.height = 5}
ggplot(plotDt, aes(x = condition, y = refFrac, fill = preference)) +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  geom_boxplot(outlier.size = 0.3, position = position_dodge(width = 0.8)) +
  facet_wrap(~ tf) +
  scale_fill_manual(values = c("Alternative" = "#1B9E77",
                               "No preference" = "#D95F02",
                               "Reference" = "#7570B3"), name = NULL) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = NULL, y = "Proportion of reads on the reference allele",
       title = "Predicted binding preference vs allelic accessibility") +
  theme(legend.position = "bottom")
```

The prediction is a separation between the two coloured boxes: sites where the motif prefers the reference allele should sit above 0.5, sites where it prefers the alternative below it. A Mann-Whitney test on those two groups puts a number on the gap, computed per factor and per condition.

```{r tfpvals, eval = runDemo, warning = FALSE}
plotDt[preference %in% c("Reference", "Alternative"),
       .(p = tryCatch(wilcox.test(refFrac ~ droplevels(preference))$p.value,
                      error = function(e) NA_real_)),
       by = .(tf, condition)][order(p)]
```

`droplevels()` matters here: `preference` is a three-level factor, and a formula-interface `wilcox.test` refuses a grouping factor that still carries an unused third level, so without it every p-value comes back as `NA` through the `tryCatch`.

# Cross-referencing published calls

The pipeline can be checked against the allele-specific variants reported in Calderon et al., *Nature Genetics* 2019 (doi:10.1038/s41588-019-0505-9), whose supplementary table lists genotypes, WASP-corrected counts and allele fractions for GWAS-overlapping ASC sites. Lifting those coordinates to the working assembly and matching per donor and site answers three questions at once: whether the direction of imbalance is reproduced, whether the reference fraction at matched sites is skewed relative to theirs, and whether the significant set as a whole sits at a comparable allele fraction. 

# Session information

```{r sessioninfo}
sessionInfo()
```
